---
title: "PANORAMIC tutorial"
vignette: >
  %\VignetteIndexEntry{PANORAMIC tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

PANORAMIC is designed for *multi-sample* spatial colocalization
analysis. It estimates sample-level spatial effects, then pools them
with multilevel meta-analysis to test group-level differences.

The published PANORAMIC paper describes the default workflow as
edge-corrected neighborhood enrichment combined with spatial bootstrap
uncertainty estimation and multilevel random-effects meta-analysis
(Chang et al., *Bioinformatics* 42(8): btag546,
https://doi.org/10.1093/bioinformatics/btag546).

Compared with single-sample analyses, PANORAMIC gives you:

-   uncertainty-aware pooling across samples
-   explicit between-sample / between-patient heterogeneity modeling
-   direct two-group contrasts (`beta_diff`, `p_diff`, `fdr_diff`)
-   consistent table and plotting utilities for downstream
    interpretation

This vignette focuses on practical usage:

1.  Build a toy multi-sample dataset.
2.  Run PANORAMIC stepwise (prepare -\> spatial stats -\> pooling).
3.  Run the one-call workflow helper (`panoramic_analyze()`).
4.  Interpret outputs and generate publication-style plots.

# Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("panoramic")
```

# Load packages

```{r libraries, message=FALSE, warning=FALSE}
library(panoramic)
library(SummarizedExperiment)
library(BiocParallel)
library(dplyr)
library(ggplot2)
```

# Workflow At A Glance

| Step        | Function                   | Primary output                                                    | Why it matters                           |
|:------------|:---------------------------|:------------------------------------------------------------------|:-----------------------------------------|
| 1           | `panoramic_prepare()`      | list of prepared `SpatialExperiment`s with cached spatial objects | standardizes per-sample inputs           |
| 2           | `panoramic_spatialstats()` | `SummarizedExperiment` with sample-level `yi` / `vi`              | quantifies within-sample spatial effects |
| 3           | `panoramic_meta_mv()`      | pooled multilevel results + differential contrasts                | tests group-level biology                |
| Convenience | `panoramic_analyze()`      | list with `prep`, `stats`, `pooled`, `tables`                     | one-call reproducible pipeline           |

# Simulate Example Data

We simulate two groups of samples with different spatial structure:

-   `group1`: mostly random cell mixing
-   `group2`: stronger local colocalization for early cell types

```{r simulate_data, message=FALSE, warning=FALSE}
set.seed(123)
toy <- panoramic_simulate_dataset(
  n_group1 = 3,
  n_group2 = 3,
  n_cells_group1 = 200,
  n_cells_group2 = 350,
  group_labels = c("group1", "group2"),
  scenario_group1 = "random",
  scenario_group2 = "colocalized",
  seed = 123
)

spe_list <- toy$spe_list
design <- toy$design
design
```

Quick geometric sanity check:

```{r plot_samples, message=FALSE, warning=FALSE}
plot_df <- do.call(rbind, lapply(spe_list, function(spe) {
  coords <- SpatialExperiment::spatialCoords(spe)
  data.frame(
    x = coords[, 1],
    y = coords[, 2],
    cell_type = SummarizedExperiment::colData(spe)$cell_type,
    sample_id = SummarizedExperiment::colData(spe)$sample_id,
    stringsAsFactors = FALSE
  )
}))

ggplot(plot_df, aes(x = x, y = y, color = cell_type)) +
  geom_point(size = 0.7, alpha = 0.7) +
  facet_wrap(~ sample_id, nrow = 2) +
  coord_equal() +
  theme_bw() +
  labs(x = "x", y = "y", color = "Cell type")
```

# Step 1: Prepare Samples

```{r prepare, message=FALSE, warning=FALSE}
prep <- panoramic_prepare(
  spe_list = spe_list,
  design = design,
  cell_type = "cell_type",
  min_cells = 5,
  window = "concave"
)
names(S4Vectors::metadata(prep[[1]])$panoramic)
```

# Step 2: Compute Sample-Level Spatial Effects

The default PANORAMIC statistic is `stat = "local_comp_enrichment"`,
which reports edge-corrected neighborhood enrichment on a percentage-point
scale relative to a sample-specific null expectation.

For tissue with strong large-scale density variation, an alternative is
`stat = "local_comp_global_enrichment"`. It compares local composition with
the sample-wide target proportion, corresponding to a random-label null that
conditions on the observed cell locations. This can be useful when the
scientific question is label mixing within an observed tissue architecture;
it should not be interpreted as correcting for spatially varying abundance or
as evidence of a direct cell-cell interaction.

```{r spatial_stats, message=FALSE, warning=FALSE}
radii_um <- c(25)

se <- panoramic_spatialstats(
  prep = prep,
  radii_um = radii_um,
  stat = "local_comp_enrichment",
  nsim = 30,
  seed = 123,
  BPPARAM = BiocParallel::SerialParam()
)
dim(se)
head(as.data.frame(rowData(se))[, c("ct1", "ct2", "radius_um", "stat")])
```

# Step 3: Pool Across Samples And Test Group Differences

`panoramic_meta_mv()` pools sample-level effects and computes
differential contrasts between groups.

```{r meta_mv, message=FALSE, warning=FALSE}
# Here patient_col and sample_col are both "sample" in toy data.
# In real data, patient_col can differ from sample_col (e.g., multiple cores per patient).

se_meta <- panoramic_meta_mv(
  se,
  patient_col = "sample",
  group_col = "group",
  sample_col = "sample",
  tau_structure = "patient",
  method = "REML",
  group_tau2 = "separate"
)
```

# Interpret Pooled And Differential Outputs

Each row in `rowData(se_meta)` is one feature (`ct1`, `ct2`,
`radius_um`) with pooled effects and contrasts.

For differential terms:

-   positive `beta_diff`: stronger spatial effect in `group2` vs
    `group1`
-   negative `beta_diff`: weaker spatial effect in `group2`
-   small `fdr_diff`: stronger evidence after multiplicity correction

```{r extract_tables, message=FALSE, warning=FALSE}
res <- panoramic_extract_contrast(se_meta)
head(res[, c("ct1", "ct2", "radius_um", "beta_diff", "p_diff", "fdr_diff")])
res %>% dplyr::filter(fdr_diff < 0.05)
```

# One-Call Workflow (`panoramic_analyze`)

Use `panoramic_analyze()` when you want a single reproducible entry
point for end-to-end runs.

```{r one_call, message=FALSE, warning=FALSE}
out <- panoramic_analyze(
  spe_list = spe_list,
  design = design,
  cell_type = "cell_type",
  radii_um = radii_um,
  nsim = 20,
  min_cells = 5,
  window = "concave",
  BPPARAM = BiocParallel::SerialParam()
)
names(out)
names(out$tables)
```

# Visualization Tools

## Forest Plot (Feature-Level)

```{r forest_plot, message=FALSE, warning=FALSE}
plot_forest(
  se_meta,
  ct1 = "A",
  ct2 = "B",
  radius_um = 25,
  group_col = "group"
)
```

## Volcano Plot (Global Differential Overview)

```{r volcano_plot, message=FALSE, warning=FALSE}
plot_volcano(se_meta)
```

## Network Summary

```{r network_plot, message=FALSE, warning=FALSE}
net <- create_spatial_network(
  se_meta,
  fdr_threshold = 0.2,
  directed = FALSE,
  leiden_resolution = 1.0
)
plot_spatial_network(
  se_meta,
  fdr_threshold = 0.2,
  directed = FALSE,
  layout = "fr",
  node_size_by = "degree"
)
```

# Practical Notes

-   Start with `local_comp_enrichment` unless you have a specific reason
    to use the global-composition or L/K-function alternatives.
-   Use at least two biological samples per group for stable contrasts.
-   Increase `nsim` for final analyses (the vignette uses modest values
    for speed).
-   Keep sample metadata (`sample`, `group`, `patient`) explicit and
    consistent early in your workflow.

# Related Resources

For complementary spatial analysis workflows, see:

-   `r BiocStyle::CRANpkg("spatstat")` family
    (`r BiocStyle::CRANpkg("spatstat.geom")`,
    `r BiocStyle::CRANpkg("spatstat.explore")`) for point-process
    spatial statistics.
-   `r BiocStyle::Biocpkg("SPIAT")` for spatial histology and tumor
    microenvironment analyses.
-   `r BiocStyle::Biocpkg("imcRtools")` for imaging mass cytometry
    spatial workflows.
-   `r BiocStyle::Biocpkg("spicyR")` for differential spatial
    localization testing in imaging data.

For methodological details and biological applications in colorectal
cancer and head and neck squamous cell carcinoma tissue microarrays, see
the published PANORAMIC paper:

-   Chang J, Espin Perez A, Molina P, Khurana R, Zhang W, Tian L,
    Plevritis SK. Multi-sample and multi-group spatial colocalization
    analysis using PANORAMIC. *Bioinformatics*. 2026;42(8):btag546.
    https://doi.org/10.1093/bioinformatics/btag546

```{r session_info}
sessionInfo()
```
